Automatically Proving Equivalence by Type-Safe Reflection

### 2017-05-16 paper-CICM-17/content/1-introduction.tex
Prune abstract and section 1

Proofs assistants like Coq~\cite{BertotC04}, Agda~\cite{DBLP:conf/tldi/Norell09} and Idris~\cite{brady2013idris} are based on Intensional Type Theories, that contain two notions of equality. One is the equality that can be manipulated in the language, called the propositional equality. This equality corresponds to the equality found in mathematics : this is a proposition that can be assumed, negated, proved or disproved. Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parameterised by the type $a$) indexed over two elements of $a$, giving $\code{Id} (a:\code{Type}):a \rightarrow a \rightarrow \code{Type}$. For
convenience, we write $(\code{Id}_a\ x\ y)$ as $(x\ =_a\ y)$.

The second notion of equality in Intentional Type Theories is a primitive concept, called judgemental equality, or definitional equality. This second equality means ``equal by definition''. The judgemental equality cannot be negated or assumed; we cannot talk about this primitive equality inside the theory.
Whether or not two expressions are equal by definition is a matter of
evaluating the definitions. For example, if $f\ :\ \mathbb{N}\ \rightarrow\
\mathbb{N}$ is defined by $f\ x\ :=\ x\ +\ 2$, then $f\ 5$ is
definitionally equal to $7$. Definitional equality entails unfolding
of functions and reductions, until no more reduction can be
performed. We denote the definitional equality by $\equiv$.

The judgemental equality has to be included in the propositional equality, because what is equal by definition should be provably equal. This is accomplished by giving constructors to the type $Id(a,a)$ and nothing when ``$a$ is not $b$".
In these theories, $Id$ is therefore implemented with the following sum type that has only one constructor :

@@

The only way for $(\code{Id}_a\ x\ y)$ to be inhabited is therefore that $x$ should be
(by definition) $y$. In this case, the constructor \code{Refl} helps to create a
proof of this equality : \code{(Refl\ x)} is precisely the proof which says that
$x=_ax$.

When reading these definitions, one could think that the propositional equality is a simple wrapper of the judgemental equality. This is not the case : the propositional equality does not only contain the judgemental equality, because in these type theories, a principle of induction is associated with each inductive type. If \code{T} is an inductive type with a constant constructor and a recursive constructor, ie, \code{T = 1 + T}, defined in Idris as :

@@

The principles of induction can be used for proving any proposition $P$, so they also work for the propositional equality. For example, we can prove that $n+0 = n$ for all $n$ by induction on the $Nat\ n$, even if $n+0 \not\equiv n$ with the usual definition of $+$, recursive on its first argument. That means that the axiom of induction has made the type $Id_{a}(x,y)$ bigger than it looks like : it contains terms that are not generated by \code{Refl} (the only constructor of $Id_{a}(x,y)$), but that are added by the inductive principles. There are therefore things which are provably equal which unfortunately are not definitionally equal after unfolding all the definitions and after full reductions. The activity of proving equalities is therefore in these theories something which isn't automatically decidable by the type-checker in the very general case.

Proving that one term is equal to another is quite common
in formal certification. This is even an everyday routine when working in a dependently typed language, and especially when intensively indexing types over values in order to capture some logical properties, as we will show in the following subsection.

@@

We revisit an example from previous work~\cite{DBLP:conf/plpv/Brady13} which shows how proof obligations may arise when a type is indexed by natural numbers. In this development, the goal is to implement a formally verified library of binary numbers. In order to ensure the correctness of the methods that will be developed, we define the types \code{Bit} and \code{Binary} as indexed over the value they represent (expressed as a natural number), following a correct-by-construction approach.

@@

We want to write a function performing the addition between two binary numbers. In order to do so, we start with an auxiliary function, which adds three bits
(the third one plays the role of a carry flag), and produces the two bits of
the result, where the first one is the most significant bit.

\begin{lstlisting}
addBit : Bit x $\rightarrow$ Bit y $\rightarrow$ Bit c $\rightarrow$ (bX ** (bY ** 
         (Bit bX, Bit bY, c + x + y = bY + 2 * bX)))
addBit b0 b0 b0 = (_ ** (_ ** (b0, b0, Refl)))
addBit b0 b0 b1 = (_ ** (_ ** (b0, b1, Refl)))
addBit b0 b1 b0 = (_ ** (_ ** (b0, b1, Refl)))
addBit b0 b1 b1 = (_ ** (_ ** (b1, b0, Refl)))
addBit b1 b0 b0 = (_ ** (_ ** (b0, b1, Refl)))
addBit b1 b0 b1 = (_ ** (_ ** (b1, b0, Refl)))
addBit b1 b1 b0 = (_ ** (_ ** (b1, b0, Refl)))
addBit b1 b1 b1 = (_ ** (_ ** (b1, b1, Refl)))
\end{lstlisting}

This function produces a dependent pair. The first component is the value \code{bX} (expressed as a natural number) representing the first bit produced (therefore of type \code{Bit\ bX}), and the second component is another dependent pair. This second dependent pair is composed of the value \code{bY} representing the second bit produced, and of a triple containing the two bits produced and a proof that these two bits correctly encode the result of the addition.
For example, on the fifth line, which corresponds to the computation $1_2 + 0_2 + 0_2 = (01)_2$, the function produces the bit \code{b0} (encoding $0_2$) and the bit \code{b1} (encoding $1_2$), and a proof that $1 + 0 + 0 = 1 + (2*0)$.

We finally define the function \code{adc} that adds two binary numbers and a carry flag. This addition works for two binary numbers expressed with the same number of bits, and it produces a result with one more bit. This result represents the sum of the three concrete values of the inputs. Therefore, we want to write :

@@

Unfortunately, this straightforward definition is rejected by Idris' type-checker because the types mismatch for both of the two patterns. The result of the first line \code{adc zero zero carry} is expected to have the type \codeL{Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)} but we provide a term of type \code{Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))}.
The problem is even worse for the second case where the expected index is \codeL{(plus\ (plus\ c\ (plus\ bit2\ (plus\ v1\ (plus\ v1\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))}
while we're trying to provide a term indexed over
\codeL{plus\ vLsb\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ 0))}.
We know that the definition of \code{adc} we have given is correct, and it has \emph{provably} the expected type, but it does not have it \emph{immediately} or \emph{by reduction} : after full reductions\footnote{In dependently typed theories, the presence of types predicated on values creates the need of using reductions also inside of types}, the expected and provided types are still different, and Idris can not figure out a proof of compatibility on its own. 
If we use provisional definitions~\cite{DBLP:conf/plpv/Brady13}, the previous definition can be accepted by the system, but we now have two proof obligations \code{adc\_lemma\_1} and \code{adc\_lemma\_2} demanding proofs of equality between the two types, and they have to be provided by hand --here with a proof script-- :

@@

As we see, this kind of proof consists of a potentially long sequence of rewriting steps, where each step uses one of the available properties: neutral element, commutativity, associativity. Without some specific automation, this sequence of rewritings must be done by the programmer.
This is time consuming and more importantly, just a small change in the left or the right hand side of the equality may invalidate the proof. That means that a minor change in our datatypes, or in the definition of \code{addBit} or \code{adc} will require us to replace this proof by another one fulfilling the same task.
These proofs become the everyday routine in any dependently-typed language, as soon as we index types over values in order to capture some logical properties.

The proof that we have written by hand only uses the existence of a neutral element, and the associativity and commutativity of $+$ on \code{Nat}. Thus, we're rewriting a term by using the properties of a commutative monoid. We would like to let a generic prover for commutative monoid find a proof on its own.

@@

Prover for some algebraic structures have already been implemented for various proof assistants, including Coq~\cite{Coq2005} and Agda\footnote{\url{http://wiki.portal.chalmers.se/agda\%5C?n=Libraries.UsingTheRingSolver}} ring prover. In this paper, we describe a certified
implementation\footnote{The implementation of our hierarchy of tactics can be found online at \url{https://github.com/FranckS/RingIdris/Provers}} of an automatic prover for equalities in a \emph{hierarchy} of algebraic
structures, including monoid, groups and rings (all potentially commutative),
for the Idris language.

The principal novelty is in the approach that we follow, using a
new kind of \emph{type-safe reflection}.  Working by reflection for implementing tactics has been done several times, including for the implementation of a ring solver for the Coq proof assistant, but without the type-safe and correct-by-construction aspects. We will
compare our approach with other implementations in section~\ref{sect:relatedWork}.

The contributions of this paper are the following :

@@

\item We present in section~\ref{sect:typeSafeReflection} a new type-safe reflection mechanism, where the reflected terms are indexed over the concrete inputs, thus providing a direct way to pull out the proofs, and providing the guarantee that the reflection of a term is indeed a faithful representation of the term.

\item The normalisation procedures are implemented by following a correct by construction approach, instead of proving the correctness afterwards with auxiliary lemmas. This approach is much more suitable for programming languages like Idris. This correct-by-construction aspect is developed in section~\ref{sect:correctByConstruction}

\item We develop a hierarchy of tactics (and not just a tactic for, say, ring) where each tactic reuses the rewriting machinery of the structure from which it naturally inherits.
For example, simplifying neutral elements is only implemented at the monoid level, and each level inheriting from it will reuse it. Reusability is difficult to obtain when we want to reuse the prover of a less expressive structure. For example, reusing the monoid prover for building the
group prover is not trivial, since we lose the possibility to express
negations ($-x$) and subtractions ($x-y$). Some specific encodings
will be presented in section~\ref{sect:reusabilityOfTheProvers} to overcome this problem.

@@

>>>>

Proofs assistants like Coq~\cite{BertotC04} 
and programming languages like 
Agda~\cite{DBLP:conf/tldi/Norell09} and
Idris~\cite{brady2013idris} are based on Intensional Type Theories that contain
two notions of equality: propositional equality, that can be
manipulated in the language, and judgemental (or definition) equality.
Propositional equality corresponds to the mathematical notion: this is
a proposition that can be assumed, negated, proved or disproved. Since in type
theory, propositions are types~\cite{How80}, the proposition that two
elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parameterised by the type $a$) indexed over two elements of $a$, giving $\code{Id} (a:\code{Type}):a \rightarrow a \rightarrow \code{Type}$. For
convenience, we write $(\code{Id}_a\ x\ y)$ as $(x\ =_a\ y)$.

Judgemental equality, on the other hand, is a primitive concept of
the type theory.
Whether or not two expressions are judgementally equal is a matter of
evaluating the definitions. For example, if $f\ :\ \mathbb{N}\ \rightarrow\
\mathbb{N}$ is defined by $f\ x\ :=\ x\ +\ 2$, then $f\ 5$ is
definitionally equal to $7$. Definitional equality entails unfolding
of functions and reductions, until no more reduction can be
performed. We denote the definitional equality by $\equiv$.

Judgemental equality is included in propositional equality
because what is equal by definition is provably equal. This is
accomplished by giving a constructor for the type $Id(a,a)$ and no
constructor when ``$a$ is not $b$".  In these theories, $Id$ is therefore
implemented with the following type with one constructor:

@@

The only way for $(\code{Id}_a\ x\ y)$ to be inhabited is therefore that $x$
equals $y$. In this case, the constructor \code{Refl} helps
to create a proof of this equality: \code{(Refl\ x)} is precisely the proof
which says that $x=_ax$.

The propositional equality does not only contain the judgemental equality,
however, because a principle of induction is associated with each inductive
type. If \code{T} is an inductive type with a constant constructor and a
recursive constructor, ie, \code{T = 1 + T}, defined in Idris as:

@@

For example, we can prove that $n+0 = n$ for all $n$ by induction on the $Nat\
n$, even if $n+0 \not\equiv n$ with the usual definition of $+$, recursive on
its first argument. So, the axiom of induction means the type $Id_{a}(x,y)$
contains not only the canonical form \code{Refl}, but also those added by
inductive principles.  There are therefore things which are \emph{provably}
equal, but not \emph{definitionally} equal.  Proving equalities is therefore in
these theories something which isn't automatically decidable by the
type-checker in the general case.

@@

Proving that one term is equal to another is common in formal verification, and
proof obligations arise naturally in dependently typed programming when
indexing types over values in order to capture some logical properties.
To demonstrate this,
we revisit an example from previous work~\cite{DBLP:conf/plpv/Brady13} which
shows how proof obligations arise when a type is indexed by natural
numbers. Our goal is to implement a verified
library of binary numbers. To ensure functional correctness,
we define types \code{Bit} and \code{Binary},
indexed over the value they represent (expressed as a natural number):

@@

We will write a function to add two binary numbers.
To do so, we begin with an auxiliary function, which adds three bits
(the third is a carry bit), and produces the two bits of
the result, where the first is the more significant bit:

\begin{lstlisting}
addBit : Bit x $\rightarrow$ Bit y $\rightarrow$ Bit c $\rightarrow$ (bX ** (bY ** 
         (Bit bX, Bit bY, c + x + y = bY + 2 * bX)))
addBit b0 b0 b0 = (_ ** (_ ** (b0, b0, Refl)))
addBit b0 b0 b1 = (_ ** (_ ** (b0, b1, Refl)))
{- ... remaining cases follow the same pattern ... -}
\end{lstlisting}

The syntax \code{(n ** t)} denotes a \emph{dependent pair}, where the 
type of the second argument \texttt{t} can refer to the first argument
\code{n}. So, we can read this type as: ``there exists a number \code{bX},
and a number \code{bY}, such that we have two bits \code{Bit bX} and
\code{Bit bY} and the sum of the input bits \code{c}, \code{x} and
\code{y} equals \code{bY + 2 * bX}.''
For example, on the second line,
which corresponds to the computation $0_2 + 0_2 + 1_2 = (01)_2$,
the function produces this bits \code{b0} and \code{b1}, and
a proof that $0 + 0 + 1 = 1 + (2 \times 0)$.

We then define the function \code{adc} that adds two binary numbers and a carry
bit. This works for two binary numbers with the same number
of bits, and produces a result with one more bit. We would like to write:

@@

Unfortunately, this definition is rejected because the types of each side do
not match.  The result of the first line \code{adc zero zero carry} is expected
to have the type:

\codeL{Binary\ 1\ ((c\ + 0)\ + 0)} but we provide a
term of type:

\code{Binary\ 1\ (c\ + (0\ + (0\ + 0)))}.

The problem is even worse for the second case where the expected index is:

\codeL{((c\ + (bit2\ + (v1\ + (v1\ + 0)))) + (bit\ + (v\ + (v\ + 0))))}
while we're trying to provide a term indexed over

\codeL{vLsb\ + (((vCarry0\ + v1) + v)\ + (((vCarry0\ + v1)\ + v)\ + 0))}.

The definition of \code{adc} we have given would behave correctly, and it has
\emph{provably} the expected type, but it does not have it \emph{immediately}
or \emph{judgementally}: after full reductions the expected and provided types are still
different.
If we use provisional definitions~\cite{DBLP:conf/plpv/Brady13}, the previous
definition can be accepted by the system, but we now have two proof obligations
\code{adc\_lemma\_1} and \code{adc\_lemma\_2} demanding proofs of equality
between the two types, and they have to be provided by hand.
For example, using a proof script:

@@

As we see, this kind of proof consists of a potentially long sequence of
rewriting steps, each using one of the available properties: neutral element,
commutativity, associativity. Without some automation, this sequence of
rewritings must be done by the programmer.  Not only is this time consuming,
just a small change in the definition may lead to a different proof obligation,
thus invalidating the proof. 
A minor change in the datatype, or 
the definition of \code{addBit} or \code{adc} will require us
to do a new proof, and thus, without support from the machine, these
proofs
could become the everyday routine in any dependently-typed language.

Our handwritten proof uses only the existence of a neutral
element, and the associativity and commutativity of $+$ on \code{Nat}. Thus,
we're rewriting a term by using the properties of a commutative monoid. We
would like a generic prover for commutative monoids to find a proof 
automatically.

@@

Provers for some algebraic structures have already been implemented for various
proof assistants, including Coq~\cite{Coq2005} and
Agda\footnote{\url{http://wiki.portal.chalmers.se/agda\%5C?n=Libraries.UsingTheRingSolver}}. 
In this paper, we describe an implementation\footnote{The
implementation of our hierarchy of tactics can be found online at
\url{https://github.com/FranckS/RingIdris/Provers}} of an automatic prover for
equalities in a \emph{hierarchy} of algebraic structures, including monoids,
groups and rings (all potentially commutative), for the Idris language.,
making the following contributions:

@@

\item We present a type-safe reflection mechanism
(section~\ref{sect:typeSafeReflection}), where the reflected terms are indexed
over the concrete terms, providing a direct way to extract proofs and
guaranteeing that the reflected term is a sound representation.

\item The normalisation procedures are implemented by following a correct by
construction approach (section~\ref{sect:correctByConstruction}), instead of proving the correctness afterwards with
auxiliary lemmas. This approach is much more suitable for programming languages
like Idris.

\item We develop a \emph{hierarchy} of tactics where each tactic reuses the
rewriting machinery of the structure from which it inherits.  For
example, simplifying neutral elements is implemented only at the monoid level,
and reused at other levels.
It is challenging
to reuse the prover of a less expressive structure; for
example, using the monoid prover to build the group prover is 
tricky because
we lose the ability to express negations ($-x$) and
subtractions ($x-y$). We present encodings
(section~\ref{sect:reusabilityOfTheProvers}) to overcome this problem.

@@

The principal novelty is in using 
\emph{type-safe reflection}.  Working by reflection for implementing tactics
has been done several times, including the implementation of a ring solver
for Coq, but without the type-safety and
correctness-by-construction. We compare our approach with other
implementations in section~\ref{sect:relatedWork}.

### 2017-05-16 paper-CICM-17/paper.tex
Prune abstract and section 1

\mainmatter              % start of the contributions
%
\title{Automatically Proving Equivalence by Type-Safe Reflection}
\subtitle{Or how to use dependent types to address some of the problems they have caused}
%
\titlerunning{xxx}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Franck Slama \and Edwin Brady}
\institute{University of St Andrews, Scotland, UK\\
\email{fs39@st-andrews.ac.uk, ecb10@st-andrews.ac.uk}}

@@

\begin{abstract}
One common difficulty with programming and reasoning with dependent types is that proof obligations arise naturally once programs become even moderately sized. For example, implementing an adder for binary numbers indexed over their natural number equivalents will naturally lead to proof obligations for equalities of expressions over universally quantified natural numbers. The need of these proofs of equalities comes, in intensional type theories, from the fact that the propositional equality enables to prove as equal terms that are not judgmentally equal, which means that the typechecker can't always obtain equalities by reduction, and therefore often needs proofs before accepting definitions.
As far as possible, we would like to solve such proof obligations automatically. In this paper, we show one way to automate these proofs by reflection for the dependently-typed programming language Idris. We will show how type-safe reflection, where reflected terms are indexed by the original Idris expression, leads to straightforward construction and manipulation of proofs. We build a hierarchy of reflexive tactics for proving equivalences in semi-groups, monoids, commutative monoids, groups, commutative groups, semi-rings and rings. We will also show how each tactic reuses the ones from the simplest structures, thus avoiding as much as possible the duplication of code and proofs.

>>>>

\mainmatter              % start of the contributions
%
\title{Automatically Proving Equivalence by Type-Safe Reflection}
%\subtitle{Or how to use dependent types to address some of the problems they have caused}
%
\titlerunning{xxx}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Franck Slama \and Edwin Brady}
\institute{University of St Andrews, Scotland, UK\\
\email{fs39@st-andrews.ac.uk, ecb10@st-andrews.ac.uk}}

@@

\begin{abstract}
One difficulty with programming with dependent types is that proof obligations
arise naturally once programs become even moderately sized. For example,
implementing an adder for binary numbers indexed over their natural number
equivalents naturally leads to proof obligations for equalities of expressions
over natural numbers. The need for these equality proofs comes, in intensional
type theories, from the fact that the propositional equality enables us to
prove as equal terms that are not judgementally equal, which means that the
typechecker can't always obtain equalities by reduction.  As far as possible,
we would like to solve such proof obligations automatically. In this paper, we
show one way to automate these proofs by reflection in the dependently typed
programming language Idris.  We show how defining reflected terms indexed by
the original Idris expression allows us to construct and manipulate proofs. We
build a hierarchy of tactics for proving equivalences in semi-groups, monoids,
commutative monoids, groups, commutative groups, semi-rings and rings. We also
show how each tactic reuses those from simpler structures, thus avoiding
duplication of code and proofs.
